---
title: "Details about NKDE"
author: "Jeremy Gelb"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 5
    fig_height: 5
    toc: true
    toc_depth: 2
    df_print: "tibble"
vignette: >
  %\VignetteIndexEntry{NKDEdetailed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: spnetwork_bilbio.bibtex
link-citations: yes
colorlinks: yes
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)

knitr::knit_hooks$set()

```

# Introduction

The goal of this vignette is to present the differences between the three Network Kernel Density Estimation methods implemented in the package **spNetwork**.

For a more detailed description, the reader should (must) read the provided references.

Let us start with the definition of a simple situation with one event and a small network.

```{r message=FALSE, warning=FALSE, fig.height = 3, fig.width  = 3}
library(spNetwork)
library(rgeos)
library(sp)

# start with de definition of some lines
wkt_lines <- c(
  "LINESTRING (0 0, 5 0)",
  "LINESTRING (0 -5, 5 -5)",
  "LINESTRING (5 0, 5 5)",
  "LINESTRING (5 -5, 5 -10)",
  "LINESTRING (5 0, 5 -5)",
  "LINESTRING (5 0, 10 0)",
  "LINESTRING (5 -5, 10 -5)",
  "LINESTRING (10 0, 10 -5)",
  "LINESTRING (10 -10, 10 -5)",
  "LINESTRING (15 -5, 10 -5)",
  "LINESTRING (10 0, 15 0)",
  "LINESTRING (10 0, 10 5)")

linesdf <- data.frame(wkt = wkt_lines,
                      id = paste("l",1:length(wkt_lines),sep=""))

geoms <- do.call(rbind,lapply(1:nrow(linesdf),function(i){
      txt <- as.character(linesdf[i,]$wkt)
      geom <- rgeos::readWKT(txt,id=i)
      return(geom)
    }))

all_lines <- SpatialLinesDataFrame(geoms, linesdf,match.ID = F)

# and the definition of one event
event <- data.frame(x=c(5),
                    y=c(-2.5))
coordinates(event) <- cbind(event$x,event$y)

# and map the situation
par(mar=c(0.1,0.1,0.1,0.1))
plot(all_lines)
plot(event,add=T,col="red",pch = 19)
```


# The simple method

The first method was proposed by @xie2008kernel. The differences with a classical KDE are: 

* the events are snapped on a network
* the distances between sampling points and events are network distance (ie. not euclidean distance)

To get a good grasp on it, we calculate here the density values and plot it in 3D. 
We sample the kernel values at each one cm on the lines.


```{r message=FALSE, warning=FALSE}
samples_pts <- lines_points_along(all_lines,0.01)

simple_kernel <- nkde(all_lines, event, w = 1,
                  samples = samples_pts, kernel_name = "quartic", 
                  bw = 10, method = "simple", div = "bw", 
                  digits = 0, tol = 0.01, grid_shape = c(1,1),
                  verbose = FALSE)
```


Let us see this in 3D !
```{r message=FALSE, warning=FALSE}
library(plot3D)

zfactor <- 1000

par(mar=c(0.1,0.1,0.1,0.1))

# step1 : plot the lines
lines_coords <- coordinates(all_lines)
x0 <- c()
y0 <- c()
x1 <- c()
y1 <- c()
for(i in 1:length(lines_coords)){
  ci <- lines_coords[[i]][[1]]
  x0 <- c(x0,ci[1,1])
  y0 <- c(y0,ci[1,2])
  x1 <- c(x1,ci[2,1])
  y1 <- c(y1,ci[2,2])
}
z0 <- rep(0,length(x0))
z1 <- rep(0,length(x1))
segments3D(x0,y0,z0,x1,y1,z1,
           zlim = c(0,20),
           phi = 15,
           theta = 30,
           axes = FALSE,
           border = NA,
           box = FALSE,
           r = 0
           )

# step2 : add the event
coords_events <- coordinates(event)
scatter3D(x = coords_events[,1],
          y = coords_events[,2],
          z = rep(0,nrow(event)),
          col="red",
          add=T, pch = 19)

# step3 : add the samples
coords_samples <- coordinates(samples_pts)
scatter3D(x = coords_samples[,1],
          y = coords_samples[,2],
          z = simple_kernel * zfactor,
          col="blue",
          cex=0.1,
          pch = 19,
          add=T)
```


As one can see, NKDE is only determined by the distance between the sampling points and the event. This is a problem, because at intersections, the event' mass is multiplied by the number of edges adjacent to that intersection. In more mathematical words, this is not a real kernel because it does not integrate to 1 on its domain.

This method remains useful for two reasons: 

* For quick data visualization. With big datasets, it might be useful to use this simple method to do a primary investigation.
* In a purely geographical view, this method is intuitive. In the case of crime analysis for example, one could argue that the *strength* of the event should not be affected by intersections on the network. In that case, the kernel function is seen as a distance decaying function.

**What to retain about the simple method**

|   Pros              |           Cons             |
|:-------------------:|:--------------------------:|
|quick to calculate   | biased (not a true kernel) |
|intuitive            | overestimate the densities |
|continuous           |                            |


# The discontinuous method

To overcome the limits of the previous method, @okabe2012spatial have proposed the discontinuous NKDE, which has been extended by @sugihara2010simple for cases where the network has cycles shorter than the bandwidth.

```{r message=FALSE, warning=FALSE}
discontinuous_kernel <- nkde(all_lines, event, w = 1,
                  samples = samples_pts, kernel_name = "quartic", 
                  bw = 10, method = "discontinuous", div = "bw", 
                  digits = 0, tol = 0.01, grid_shape = c(1,1),
                  verbose = FALSE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plot3D)

zfactor <- 1000
a1 <- 1
a2 <- 0

par(mar=c(0.1,0.1,0.1,0.1))

##step1 : plot the lines
lines_coords <- coordinates(all_lines)
x0 <- c()
y0 <- c()
x1 <- c()
y1 <- c()
for(i in 1:length(lines_coords)){
  ci <- lines_coords[[i]][[1]]
  x0 <- c(x0,ci[1,1])
  y0 <- c(y0,ci[1,2])
  x1 <- c(x1,ci[2,1])
  y1 <- c(y1,ci[2,2])
}
z0 <- rep(0,length(x0))
z1 <- rep(0,length(x1))
segments3D(x0,y0,z0,x1,y1,z1,
           zlim = c(0,15),
           phi = 15,
           theta = 30,
           axes = FALSE,
           border = NA,
           box = FALSE,
           r = 0
           )

##step2 : add the event
coords_events <- coordinates(event)
scatter3D(x = coords_events[,1],
          y = coords_events[,2],
          z = rep(0,nrow(event)),
          col="red",
          add=T, pch = 19)

##step3 : add the samples
coords_samples <- coordinates(samples_pts)
scatter3D(x = coords_samples[,1],
          y = coords_samples[,2],
          z = discontinuous_kernel * zfactor,
          col="blue",
          cex=0.1,
          pch = 19,
          add=T)
```


As one can see, the values of the NKDE are split at intersections to avoid the multiplication of the mass observed in the simple version. However, this leads to a discontinuous NKDE, which is counter-intuitive. This leads to sharp differences between density values in the network, and could be problematic in networks with a lot of intersections.

**What to retain about the discontinuous method**

|   Pros              |           Cons             |
|:-------------------:|:--------------------------:|
|quick to calculate   | counter-intuitive         |
|unbiased             | discontinuous              |
|                     |   contrasted results       |


# The continuous method

Finally, the continuous method tries to take the best of the two worlds: it adjusts the values of the NKDE at intersections to ensure that it integrates to 1 on its domain, and apply a backward correction to force the density values to be continuous. This process is accomplished by a recursive function, described in the book *Spatial Analysis Along Networks* [@okabe2012spatial]. This function is very time consuming, so it might be necessary to stop it when the recursion is too deep. Considering that the kernel density is divided at each intersection, stopping the function at deep level 15 would yield results very close to the true values.


```{r message=FALSE, warning=FALSE}
samples_pts <- lines_points_along(all_lines,0.01)

continuous_kernel <- nkde(all_lines, event, w = 1,
                  samples = samples_pts, kernel_name = "quartic", 
                  bw = 10, method = "continuous", div = "bw", 
                  digits = 0, tol = 0.01, grid_shape = c(1,1),
                  verbose = FALSE, max_depth = 15)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plot3D)

zfactor <- 1000

par(mar=c(0.1,0.1,0.1,0.1))

##step1 : plot the lines
lines_coords <- coordinates(all_lines)
x0 <- c()
y0 <- c()
x1 <- c()
y1 <- c()
for(i in 1:length(lines_coords)){
  ci <- lines_coords[[i]][[1]]
  x0 <- c(x0,ci[1,1])
  y0 <- c(y0,ci[1,2])
  x1 <- c(x1,ci[2,1])
  y1 <- c(y1,ci[2,2])
}
z0 <- rep(0,length(x0))
z1 <- rep(0,length(x1))
segments3D(x0,y0,z0,x1,y1,z1,
           zlim = c(0,15),
           phi = 15,
           theta = 30,
           axes = FALSE,
           border = NA,
           box = FALSE,
           r = 0
           )

##step2 : add the event
coords_events <- coordinates(event)
scatter3D(x = coords_events[,1],
          y = coords_events[,2],
          z = rep(0,nrow(event)),
          col="red",
          add=T, pch = 19)

##step3 : add the samples
coords_samples <- coordinates(samples_pts)
scatter3D(x = coords_samples[,1],
          y = coords_samples[,2],
          z = continuous_kernel * zfactor,
          col="blue",
          cex=0.1,
          pch = 19,
          add=T)
```


As one can see, the values of the NKDE are continuous, and the density values close to the events have been adjusted. This leads to smoother results than the discontinuous method.

**what to retain about the discontinuous method**

|   Pros              |           Cons             |
|:-------------------:|:--------------------------:|
| unbiased            | long calculation time      |
| continuous          | smoother values            |


# References
