---
title: "Spatio-temporal dbscan with network distance"
author: "Jeremy Gelb"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal dbscan with network distance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "../spnetwork_bilbio.bibtex"
link-citations: yes
colorlinks: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
backup_option <- options()
base_wd <- getwd()
library(ggplot2)
```

# Introduction

This vignette is just a simple example about what can be done with `spNetwork` and little bit of imagination. We will performe here a spatio-temporal dbscan clustering with network distances on the `bike_accident` dataset. Here is the procedure:

1. calculating a network distance matrix between each pair of points;
2. calculating a temporal distance matrix between each pair of points;
3. Calculating a binary matrix as the intersection of the two previous matrices according to a temporal and a network maximum distance thresholds;
4. using the dbscan algorithm on the binary matrix;
5. analysing the results.

# Calculating the two distance matrix

This is the first step, we will start with the network matrix.

```{r message=FALSE, warning=FALSE}
# first load data and packages
library(sf)
library(spNetwork)
library(spdep)
library(dbscan)
library(tmap)

data(mtl_network)
data(bike_accidents)

distance_mat_listw <- network_listw(bike_accidents, mtl_network,
                                  maxdistance = 5000,
                                  dist_func = "identity",
                                  matrice_type = "I",
                                  grid_shape = c(1,1))

# spdep changed its name rom nb_list to neighbours
distance_mat_listw$neighbours <- distance_mat_listw$nb_list
distance_mat_net <- listw2mat(distance_mat_listw)
```

Great, now we will calculate a temporal distance matrix in days between the accidents.

```{r message=FALSE, warning=FALSE}
bike_accidents$dt <- as.POSIXct(bike_accidents$Date, format = "%Y/%m/%d")
start_time <- min(bike_accidents$dt)
bike_accidents$time <- difftime(bike_accidents$dt,start_time, units = "day")

temporal_mat <- as.matrix(dist(bike_accidents$time))
```

# Calculating the intersection matrix

We select here the two following thresholds: 500m and 25 days. We calculate a binary matrix indicating if two points are close enough in time and space to belong to the same cluster.

```{r message=FALSE, warning=FALSE, out.width = "60%", fig.align= "center"}
binary_mat <- as.integer(temporal_mat <= 25 & distance_mat_net<= 400)
dim(binary_mat) <- dim(temporal_mat)
```

# Applying the dbscan algorithm

The last step is to just apply the dbscan algorithm!

```{r message=FALSE, warning=FALSE}
result <- dbscan(binary_mat, eps = 1, minPts = 5)
result
```

# Analysing the results

We will first map the clusters (everybody loves map).

```{r message=FALSE, warning=FALSE}
tmap_mode("view")

bike_accidents$cluster <- as.character(result$cluster)
out_cluster <- subset(bike_accidents,bike_accidents$cluster == "0")
in_cluster <- subset(bike_accidents,bike_accidents$cluster != "0")

tm_shape(out_cluster) + 
  tm_dots("black", alpha = 0.3, size = 0.01) + 
  tm_shape(in_cluster) + 
  tm_dots("cluster")


```

And finally, we will plot the time period of each cluster.

```{r message=FALSE, warning=FALSE}

ggplot(in_cluster) + 
  geom_point(aes(x = dt, y = cluster, color = cluster)) +
  scale_x_datetime(date_labels = "%Y/%m")
```


That's all folks! I hope this short example was interesting!
